Compositionality test
========================================================

This code is meant to test the compositionality issue when performing differential analysis on relative abundances of microbes. The code was developed to test the species that significantly changed their relative abundance between time points in the paper:
"Roux-en-Y gastric bypass surgery of morbidly obese patients shows swift and persistent changes of the individual gut microbiota"

The code needs a relative abundance table for species and a grouping file as input, and produces a table with Pvalues after compositionality test for each species.

# Loading relative abundances table and grouping file
```{r data loading}
setwd(".")
rm(list = ls())

## Reading the relative abundance table:

# This table should have samples as rows and taxa as columns, otherwise transpose it
# All microbial feature selection and data transformation that you would like to apply to your data have to be performed before you read your table in this code. The code assumes that one of your taxa will be an unknow term (the HQ reads you could not map to any species)
relative.abundances<-read.table("species_table.tsv", sep='\t', header=TRUE, row.names=1)
head(relative.abundances)
dim(relative.abundances)

## Reading the grouping file:

# This file is a two-column file containing the first columns for the samples and the second for the category. This code is meant to compare two groups of paired samples. So we need two time categories.
groups<-read.table("Groups.tsv", sep='\t', header=TRUE, row.names=1)
colnames(groups)[1]<-"Group"

# Check that your samples are in the same order than your stratifying column
relative.abundances<-relative.abundances[rownames(groups),]
rownames(relative.abundances)
rownames(groups)

```

``` {r loop} 

# Initializying the fold change and P-values dataframes for the results
Pvalue.g1.g2<-data.frame(matrix(ncol = dim(relative.abundances)[2]-1, nrow = dim(relative.abundances)[2]-1))

# The for loop will go through all the species in the species table leaving one out each iteration and calculating Pvalues of a Wilcoxon test for the rest of the species:
for (i in 2:length(colnames(relative.abundances))) {
  
  print(colnames(relative.abundances)[i])
  
  # Leaving one out
  M.one.out<-relative.abundances[,-i]
  
  # Renormalizing
  M.one.out.norm<-M.one.out/rowSums(M.one.out)
  
  # Running Wilcoxon test
  compos.test<-wilcox_fc(M.one.out.norm, colnames(relative.abundances)[i])
  compos.test<-compos.test[order(compos.test[,1]),]
  Pvalue.g1.g2[,i-1]<-compos.test$Pvalue
}

M.without.unknown<-as.data.frame(relative.abundances[,-1])
colnames(Pvalue.g1.g2)<-colnames(M.without.unknown)
rownames(Pvalue.g1.g2)<-sort(colnames(M.without.unknown))

```

```{r table with the results}

# We find which is the least significant Pvalue 
pval.test.g1.g2<-data.frame(sapply(Pvalue.g1.g2[,1:dim(Pvalue.g1.g2)[2]], as.numeric))
rownames(pval.test.g1.g2)<-rownames(Pvalue.g1.g2)
pval.test.g1.g2$Least.signif.pval<-apply(pval.test.g1.g2[,1:dim(pval.test.g1.g2)[2]],1,max, na.rm=TRUE)

# We create a table to study the results after the compositionality test
P.value.after.compositionality.test<-data.frame(Taxa = rownames(pval.test.g1.g2), Pvalue =  pval.test.g1.g2$Least.signif.pval)

# Sorting
P.value.after.compositionality.test<-P.value.after.compositionality.test[with(P.value.after.compositionality.test, order(Pvalue)),]

# Writing a table with the results
write.table(P.value.after.compositionality.test, file="P.value.after.compositionality.test.tsv", quote=FALSE, sep='\t', row.names=FALSE, col.names=TRUE)

```

#To test for significant changes in the abundance of species between two time points we used Wilcoxon signed-rank test. Modify your code depending on your data and your groups to be compared.
``` {r wilcoxon function}

wilcox_fc = function (M, skip) {

# We remove the unknown term
# We convert the matrix to a dataframe
M<-as.data.frame(M[,-1])

# We add to the matrix the grouping about time points and the pairing samples info
groups$Group<-factor(groups$Group, levels=c("pre", "3 mo"), labels=c("Baseline", "3 MO"))
M.strat<-cbind(M, Group = groups$Group)

# We perform the wilcoxon signed-rank tests
g1.g2.pvals<-c()
num.taxa<-dim(M.strat)[2]-1
for (i in 1:num.taxa){
  g1.g2.pvals[i] = wilcox.test(M.strat[,i] ~ Group, M.strat, paired = TRUE)$p.value
}

# we assign names to the pvalues
names(g1.g2.pvals)<-colnames(M)

newrow<-c(skip, NA, NA)

# We create a table with the taxa and the Pvalues for the test
table<-data.frame(
  Taxa = as.character(names(g1.g2.pvals)),
  Pvalue = as.numeric(g1.g2.pvals),
  stringsAsFactors=FALSE)
table<-rbind(table, newrow)

return(table)

}
```
