Compositionality test
========================================================

This code was developed to test if our results of differential analysis on relative abundances of microbes were real or affected by compositional effects. The compositionality issue can arise basically when an isolated increase in absolute abundance of just one taxon will lead to dissipated decrease in relative abundance of all other taxa as the relative abundance must always sum to 1. 

For context and references check our paper:
"Roux-en-Y gastric bypass surgery of morbidly obese patients shows swift and persistent changes of the individual gut microbiota"

The code needs as input an species relative abundance table and a grouping file, and produces a table with P-values tested for compositional issue for each species.

# Loading relative abundances and grouping file
```{r data loading}
setwd(".")
rm(list = ls())

## Reading the relative abundance table:

# This table should have samples as rows and taxa as columns, otherwise transpose it

# All microbial feature selection and data transformation that you would like to apply to your data have to be performed before you read your table in this code. The code assumes that one of your taxa will be an unknow term (the HQ reads you could not map to any species)
relative.abundances<-read.table("species_table.tsv", sep='\t', header=TRUE, row.names=1)
head(relative.abundances)
dim(relative.abundances)

## Reading the grouping file:

# This file is a two-column file containing the sample names and the groups they belong to. In this example we compared two time points.
groups<-read.table("Groups.tsv", sep='\t', header=TRUE, row.names=1)
colnames(groups)[1]<-"Group"

# Check that your samples are in the same order than your stratifying column
relative.abundances<-relative.abundances[rownames(groups),]
rownames(relative.abundances)
rownames(groups)

```

# Algorithm
``` {r loop} 

# Initializying the P-values dataframes for the results
Pvalue.g1.g2<-data.frame(matrix(ncol = dim(relative.abundances)[2]-1, nrow = dim(relative.abundances)[2]-1))

# The for loop will go through all the species in the species table leaving this species out and running a Wilcoxon test for each of the rest to know if there are significant differences on the relativs abundance of these species between the 2 groups when another of the species fully disappears
for (i in 2:length(colnames(relative.abundances))) {
  
  print(colnames(relative.abundances)[i])
  
  # Leaving one out
  M.one.out<-relative.abundances[,-i]
  
  # Renormalizing
  M.one.out.norm<-M.one.out/rowSums(M.one.out)
  
  # Running Wilcoxon test
  compos.test<-wilcox(M.one.out.norm, colnames(relative.abundances)[i])
  compos.test<-compos.test[order(compos.test[,1]),]
  Pvalue.g1.g2[,i-1]<-compos.test$Pvalue
}

# Adding the right names
M.without.unknown<-as.data.frame(relative.abundances[,-1])
colnames(Pvalue.g1.g2)<-colnames(M.without.unknown)
rownames(Pvalue.g1.g2)<-sort(colnames(M.without.unknown))

```

# Producing the results table (P-values tested for compositionality)
```{r table with the results}

# We take the least significant P-value for each species as the P-value tested for compostionality. Although it may be a bit conservative approach we consider that if only one of these P-values for each species become unsignificant the result is affected by compositionality
pval.test.g1.g2<-data.frame(sapply(Pvalue.g1.g2[,1:dim(Pvalue.g1.g2)[2]], as.numeric))
rownames(pval.test.g1.g2)<-rownames(Pvalue.g1.g2)
pval.test.g1.g2$Least.signif.pval<-apply(pval.test.g1.g2[,1:dim(pval.test.g1.g2)[2]],1,max, na.rm=TRUE)

# We create a table to study the results after the compositionality test
P.value.after.compositionality.test<-data.frame(Taxa = rownames(pval.test.g1.g2), Pvalue =  pval.test.g1.g2$Least.signif.pval)

# Sorting
P.value.after.compositionality.test<-P.value.after.compositionality.test[with(P.value.after.compositionality.test, order(Pvalue)),]

# Writing a table with the results
write.table(P.value.after.compositionality.test, file="P.value.after.compositionality.test.tsv", quote=FALSE, sep='\t', row.names=FALSE, col.names=TRUE)

```

# Function for the statistical test
``` {r wilcoxon function}

# To test for significant changes in the relative abundance of species between two time points we used Wilcoxon signed-rank test (appropiate to compare two groups of paired samples when data is not normally distributed as it is often in the microbiome filed). Modify your code/use another statistical test depending on your data and your groups to be compared
wilcox = function (M, skip) {

# We remove the unknown term
# We convert the matrix to a dataframe
M<-as.data.frame(M[,-1])

# We add to the matrix the grouping about time points and the pairing samples info
groups$Group<-factor(groups$Group, levels=c("pre", "3 mo"), labels=c("Baseline", "3 MO"))
M.strat<-cbind(M, Group = groups$Group)

# We perform the wilcoxon signed-rank tests
g1.g2.pvals<-c()
num.taxa<-dim(M.strat)[2]-1
for (i in 1:num.taxa){
  g1.g2.pvals[i] = wilcox.test(M.strat[,i] ~ Group, M.strat, paired = TRUE)$p.value
}

# we assign names to the pvalues
names(g1.g2.pvals)<-colnames(M)

newrow<-c(skip, NA, NA)

# We create a table with the taxa and the Pvalues for the test
table<-data.frame(
  Taxa = as.character(names(g1.g2.pvals)),
  Pvalue = as.numeric(g1.g2.pvals),
  stringsAsFactors=FALSE)
table<-rbind(table, newrow)

return(table)

}
```
